---
title: "Global CO₂ Emissions Analysis"
author: "Son Nguyen"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    code_folding: hide
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 8,
  fig.height = 5
)
# Install & load packages
pkgs <- c("ggplot2","dplyr","lubridate","forcats","tidyr","scales","viridis","zoo")
for(pkg in pkgs) if(!requireNamespace(pkg, quietly=TRUE)) install.packages(pkg)
lapply(pkgs, library, character.only=TRUE)
````

# 1. Introduction

This report analyzes global CO₂ emissions data from Our World in Data (OWID). We will:

1. Fetch and clean the dataset
2. Compute per-capita and total emissions, and GDP per capita
3. Visualize global trends, country snapshots, and economic relationships
4. Fit a regression model predicting CO₂ per capita from GDP per capita

# 2. Data Loading & Preparation

```{r load-data}
url <- "https://raw.githubusercontent.com/owid/co2-data/master/owid-co2-data.csv"
co2_raw <- read.csv(url, stringsAsFactors=FALSE)

co2 <- co2_raw %>%
  filter(!is.na(co2_per_capita), !is.na(gdp), population > 0, year >= 1960) %>%
  transmute(
    country   = country,
    iso_code  = iso_code,
    date      = as.Date(paste0(year, "-01-01")),
    year      = year,
    co2_pc    = co2_per_capita,
    total_co2 = co2,                # Mt
    gdp_pc    = gdp / population
  )

latest <- co2 %>%
  group_by(iso_code, country) %>%
  slice_max(year, n=1, with_ties=FALSE) %>%
  ungroup()

global_ts <- co2 %>%
  group_by(date) %>%
  summarize(avg_co2_pc = mean(co2_pc, na.rm=TRUE), .groups="drop")
```

# 3. Global Average CO₂ per Capita Over Time

```{r global-ts}
library(ggplot2)
p1 <- ggplot(global_ts, aes(date, avg_co2_pc)) +
  geom_line(color="forestgreen") +
  labs(
    title = "Global Average CO₂ per Capita Over Time",
    x     = "Date",
    y     = "t CO₂ per Capita"
  ) +
  theme_minimal()
print(p1)
```

# 4. GDP vs CO₂ per Capita (Latest Snapshot)

```{r scatter-gdp-co2}
p2 <- ggplot(latest, aes(gdp_pc, co2_pc)) +
  geom_point(alpha=0.5) +
  geom_smooth(method="loess", se=FALSE, color="darkred") +
  labs(
    title = "GDP per Capita vs CO₂ per Capita (Latest)",
    x     = "GDP per Capita (USD)",
    y     = "t CO₂ per Capita"
  ) +
  theme_minimal()
print(p2)
```

# 5. Regression Analysis

```{r regression}
model <- lm(co2_pc ~ gdp_pc, data=latest)
cat("Linear model summary:\n")
print(summary(model))

p3 <- ggplot(latest, aes(gdp_pc, co2_pc)) +
  geom_point(alpha=0.3) +
  geom_smooth(method="lm", se=TRUE, color="blue") +
  labs(
    title    = "Regression: CO₂ per Capita ~ GDP per Capita",
    subtitle = paste0(
      "R²=", round(summary(model)$r.squared,3),
      "; p=", signif(summary(model)$coefficients[2,4],3)
    ),
    x = "GDP per Capita (USD)",
    y = "t CO₂ per Capita"
  ) +
  theme_minimal()
print(p3)
```

# 6. Top Emitters by CO₂ per Capita

```{r top10-pc}
top10_pc <- latest %>%
  arrange(desc(co2_pc)) %>%
  slice_head(n=10)
p4 <- ggplot(top10_pc, aes(fct_reorder(country, co2_pc), co2_pc)) +
  geom_col(fill="darkslateblue") +
  coord_flip() +
  labs(
    title = "Top 10 Countries by CO₂ per Capita (Latest)",
    x     = NULL,
    y     = "t CO₂ per Capita"
  ) +
  theme_minimal()
print(p4)
```

# 7. Trends for Top 5 CO₂ Per-Capita Emitters

```{r ts-top5}
top5_isos <- top10_pc$iso_code[1:5]
ts_top5 <- co2 %>% filter(iso_code %in% top5_isos)
p5 <- ggplot(ts_top5, aes(date, co2_pc, colour=country)) +
  geom_line() +
  labs(
    title = "CO₂ per Capita Over Time: Top 5 Emitters",
    x     = "Date",
    y     = "t CO₂ per Capita",
    colour = "Country"
  ) +
  theme_minimal()
print(p5)
```

# 8. CO₂ per Capita by GDP Quartile

```{r boxplot-quartiles}
latest_q <- latest %>%
  mutate(gdp_q = ntile(gdp_pc, 4))
p6 <- ggplot(latest_q, aes(factor(gdp_q), co2_pc, fill=factor(gdp_q))) +
  geom_boxplot() +
  scale_fill_viridis_d(name="GDP Quartile") +
  labs(
    title = "CO₂ per Capita by GDP per Capita Quartile",
    x     = "GDP Quartile",
    y     = "t CO₂ per Capita"
  ) +
  theme_minimal()
print(p6)
```

# 9. Violin Plot by GDP Quartile

```{r violin-quartiles}
p7 <- ggplot(latest_q, aes(factor(gdp_q), co2_pc, fill=factor(gdp_q))) +
  geom_violin() +
  scale_fill_viridis_d(option="turbo", name="GDP Quartile") +
  labs(
    title = "Distribution of CO₂ per Capita by GDP Quartile",
    x     = "GDP Quartile",
    y     = "t CO₂ per Capita"
  ) +
  theme_minimal()
print(p7)
```

# 10. Heatmap: CO₂ per Capita by Year & GDP Quartile

```{r heatmap}
heat_data <- co2 %>%
  inner_join(latest_q %>% select(iso_code, gdp_q), by="iso_code") %>%
  group_by(year = lubridate::year(date), gdp_q) %>%
  summarize(avg_pc = mean(co2_pc, na.rm=TRUE), .groups="drop")
p8 <- ggplot(heat_data, aes(year, factor(gdp_q), fill=avg_pc)) +
  geom_tile(color="white") +
  scale_fill_viridis(name="Avg t CO₂/capita") +
  labs(
    title = "Avg CO₂ per Capita by Year & GDP Quartile",
    x     = "Year",
    y     = "GDP Quartile"
  ) +
  theme_minimal()
print(p8)
```

# 11. Cumulative Sum of CO₂ per Capita

```{r cumulative-sum}
cum_pc <- co2 %>%
  arrange(date) %>%
  group_by(date) %>%
  summarize(sum_pc = sum(co2_pc, na.rm=TRUE), .groups="drop") %>%
  mutate(cum_pc = cumsum(sum_pc))
p9 <- ggplot(cum_pc, aes(date, cum_pc)) +
  geom_line(color="purple") +
  labs(
    title = "Cumulative Sum of CO₂ per Capita Over Time",
    x     = "Date",
    y     = "Cumulative t CO₂ per Capita"
  ) +
  theme_minimal()
print(p9)
```

# 12. Year-over-Year % Change in Global Avg CO₂ per Capita

```{r yoy-change}
pct_change <- global_ts %>%
  arrange(date) %>%
  mutate(pct = 100*(avg_co2_pc - lag(avg_co2_pc))/lag(avg_co2_pc))
p10 <- ggplot(pct_change, aes(date, pct)) +
  geom_line(color="orange") +
  labs(
    title = "Year-over-Year % Change in Global Avg CO₂ per Capita",
    x     = "Date",
    y     = "% Change"
  ) +
  theme_minimal()
print(p10)
```

# 13. CO₂ per Capita vs Total CO₂ Emissions

```{r scatter-total}
library(scales)
p11 <- ggplot(latest, aes(total_co2, co2_pc)) +
  geom_point(alpha=0.5) +
  geom_smooth(method="lm", se=FALSE, color="darkgreen") +
  scale_x_log10(labels=comma) +
  labs(
    title = "CO₂ per Capita vs Total CO₂ Emissions (Latest)",
    x     = "Total CO₂ (Mt, log scale)",
    y     = "t CO₂ per Capita"
  ) +
  theme_minimal()
print(p11)
```

# 14. Conclusions

This analysis reveals:

* Global per‐capita CO₂ emissions have trended upward since 1960.
* Wealthier countries tend to emit more CO₂ per person (strong positive relationship).
* The top 10 per‐capita emitters are small but high‐emission nations.
* GDP quartile analysis highlights distributional differences.
* Regression confirms GDP per capita is a significant predictor of emissions per capita.

Further work could segment by region, explore non‐linear models, or incorporate renewable energy metrics.
